Inference for multivariate peaks over threshold models

### 2021-11-30 abstract.tex
updates

{\bf We need an abstract}
Extreme analysis

Drawing on the constructive definition for the multivariate Pareto introduced in~\cite{ferreira2014} which separates the radial component of the RV from the angular, we introduce a method for inferring the distribution of the angular component.

>>>>

We consider a constructive definition of the multivariate Pareto that factorizes the 
random vector into a radial component and an independent angular component. The former
follows a univariate Pareto distribution. The latter is defined on the surface of the 
positive orthant of the unit  hypercube.  In this paper, we  propose a method for
inferring the distribution of the angular component.  We   identify its support 
as the limit of the positive orthant of the unit $p$--norm spheres, 
and describe a method to project a distribution onto those spheres. To define a 
distribution in the $p$--norm spheres, we introduce a family of projected Gamma 
distributions. This family serves as a building block for a flexible family of 
distributions obtained as  a Dirichlet process mixture of projected Gammas. For 
model comparison and assessment, we discuss model scoring methods appropriate 
to distributions on the unit hypercube.  In particular, we develop a kernel metric 
that produces a proper scoring rule.   We present a simulation study to compare 
different modeling choices using the proposed  scoring rules. We then apply our 
approach to study the extreme values of the  magnitude of the integrated vapor 
transport (IVT), that describes the rate of flow of moisture in the atmosphere 
along the coast of California for the years of 1979 through 2020.

% EOF

### 2021-11-30 evaluation.tex
updates

To reduce the computations needed to calculate the energy score we define a kernel as {\bf Please write a clearer description here.} our kernel an upper bound on distance between points on separate faces as the length of the path from the source point, to an optimal point on the intersection between faces, to the destination point.
\begin{prop}\label{prop:g}
Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, for $\ell, \jmath \in \{1, \ldots , d\}$. Define $g$ as
\[  
    g(\bm{a},\bm{b}) = \min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
\]
Then $g$  is negative definite kernel.
\end{prop}

{\em Proof:}
When $\bm{a}$ and $\bm{b}$ are on the same face, then $\ell = \jmath$, and $g$ is the Euclidean distance between $\bm{a}$ and $\bm{b}$, which is a  negative definite kernel.  When $\bm{a}$ and $\bm{b}$ are on separate
  faces, let 
  \[\bm{d} = 
    \argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
  \]
  We have that $g(\bm{a}, \bm{b}) = g(\bm{b}, \bm{a})$ due to the uniqueness of $\bm{d}$ and the symmetry of the Euclidean distance. Additionally, given $n$ and  a collection $\alpha_1,\ldots,\alpha_n\in{\mathbb R}$ such that  $\sum_{i = 1}^n\alpha_i = 0$, then, 
  \begin{equation*}
    \begin{aligned}
      \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2) &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{d} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{d}}{2}\bigg]\\
      &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{d} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{d}}{2} \leq0,
    \end{aligned}
  \end{equation*}
  as the Euclidean norm is a negative definite kernel.$\hfill\Box$

Evaluating $g$ as defined in Proposition \ref{prop:g} requires the solution of a $(d-2)$-dimensional 
  optimization problem.  The following proposition provides a straightforward solution.
\begin{prop}
    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, for $\ell, \jmath \in \{1, \ldots , d\}$.  For $\ell\neq \jmath$, the 
  transformation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the $\ell$th axis 
  produces the vector $\bm{b}^\prime$, where
  \begin{equation}
    \label{eqn:rotation}
     \bm{b}^{\prime}_i = P_{\jmath\ell}(\bm{b}) = 
    \begin{cases}
        b_{i} &\text{for }i\neq \jmath,\ell\\
        1 &\text{for }i = \ell\\
        2 - b_{\ell} &\text{for }i = \jmath
    \end{cases}.
  \end{equation}
  Then $g(\bm{a},\bm{b}) = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}$.
\end{prop}
{\em Proof:}
Notice that for 
  $\bm{c} \in {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$. {\bf This is true just because you do the math, the invariance of Euclidean norms is not needed.} Then, recalling the definition of $\bm{d}$ in the proof of Proposition \ref{prop:g}, we have that
  \begin{equation*}
    \begin{aligned}
    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{d}}{2} + \pnorm{\bm{b} - \bm{d}}{2}\\
    &= \pnorm{\bm{a} - \bm{d}}{2} + \pnorm{\bm{b}^{\prime} - \bm{d}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
    \end{aligned}
  \end{equation*}
  The last equality is due to the fact that $\bm{a}$ and $\bm{b}'$ belong to the same hyperplane.
  $\hfill\Box$

>>>>

To reduce the computations needed to calculate the energy score we define a kernel as 
in the following proposition.
\begin{prop}\label{prop:g}
Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, for $\ell, \jmath \in \{1, \ldots , d\}$. Define $g$ as
\[  
    g(\bm{a},\bm{b}) = \min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
\]
Then $g$  is negative definite kernel.
\end{prop}

{\em Proof:}
When $\bm{a}$ and $\bm{b}$ are on the same face, then $\ell = \jmath$, and $g$ is the Euclidean distance between $\bm{a}$ and $\bm{b}$, which is a  negative definite kernel.  When $\bm{a}$ and $\bm{b}$ are on separate
  faces, let 
  \[\bm{e} = 
    \argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
  \]
  We have that $g(\bm{a}, \bm{b}) = g(\bm{b}, \bm{a})$ due to the uniqueness of $\bm{e}$ and the symmetry of the Euclidean distance. Additionally, given $n$ and  a collection $\alpha_1,\ldots,\alpha_n\in{\mathbb R}$ such that  $\sum_{i = 1}^n\alpha_i = 0$, then, 
  \begin{equation*}
    \begin{aligned}
      \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2) &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{e} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{e}}{2}\bigg]\\
      &= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{e} - \bm{x}_1}{2} + \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{e}}{2} \leq0,
    \end{aligned}
  \end{equation*}
  as the Euclidean norm is a negative definite kernel.$\hfill\Box$

Evaluating $g$ as defined in Proposition \ref{prop:g} requires the solution of a $(d-2)$-dimensional 
  optimization problem.  The following proposition provides a straightforward solution.
\begin{prop}
    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, for $\ell, \jmath \in \{1, \ldots , d\}$.  For $\ell\neq \jmath$, the 
  transformation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the $\ell$th axis 
  produces the vector $\bm{b}^\prime$, where
  \begin{equation}
    \label{eqn:rotation}
     \bm{b}^{\prime}_i = P_{\jmath\ell}(\bm{b}) = 
    \begin{cases}
        b_{i} &\text{for }i\neq \jmath,\ell\\
        1 &\text{for }i = \ell\\
        2 - b_{\ell} &\text{for }i = \jmath
    \end{cases}.
  \end{equation}
  Then $g(\bm{a},\bm{b}) = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}$.
\end{prop}
{\em Proof:}
Notice that for 
  $\bm{c} \in {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$. Then, recalling the definition of 
     $\bm{e}$ in the proof of Proposition \ref{prop:g}, we have that
  \begin{equation*}
    \begin{aligned}
    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b} - \bm{e}}{2}\\
    &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b}^{\prime} - \bm{e}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
    \end{aligned}
  \end{equation*}
  The last equality is due to the fact that $\bm{a}$ and $\bm{b}'$ belong to the same hyperplane.
  $\hfill\Box$

### 2021-11-30 methodology_projgamma.tex
updates

@@

The second form of centering distribution we explore is a multivariate log-normal distribution on the shape parameters $\bm{\alpha}_j$, with independent Gamma $\beta_{j\ell}$ rate parameters.  
\begin{equation}
    G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right) = \mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right).
\end{equation}
This model is completed with a normal prior on $\bm{\mu}$, an inverse Wishart prior on $\bm{\Sigma}$, and Gamma priors on $\zeta_{\ell}$ and $\sigma_{\ell}$, and $\eta$.  This model is denoted as the \emph{projected Gamma--log-normal} (PG--LN) model.  We also explore a restricted Gamma form of this model as above, where $\beta_{\ell} = 1$ for all $\ell$.  This is denoted as the \emph{projected restricted Gamma--log-normal} (PRG--LN) model.  In the interest of brevity, we omit the inferential forms for this model, sufficing to say that the Gamma prior for $\beta_{j\ell}$ means it can again be integrated out, improving mixing on sampling $\bm{\alpha}_j$ with a joint Metropolis step; the normal prior for $\mu$ is conjugate for the log-normal $\log\bm{\alpha}_j$, and updated via a Gibbs step; and an inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, implying that it can also be updated via a Gibbs step.

>>>>

\begin{figure}[ht]
    \centering
  \begin{subfigure}[b]{0.47\textwidth}
    \includegraphics[width=\textwidth]{./images/p_sphere}
    \caption{${\mathbb S}_p^1$ for $p = 1,\ldots, 10$\label{fig:psphere}}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.47\textwidth}
    \includegraphics[width=\textwidth]{./images/p_project}
    \caption{Projection of data on ${\mathbb S}_{\infty}^1$ onto ${\mathbb S}_p^1$\label{fig:pproject}}
  \end{subfigure}
\end{figure}

@@

The second form of centering distribution we explore is a multivariate log-normal distribution on the shape parameters $\bm{\alpha}_j$, with independent Gamma $\beta_{j\ell}$ rate parameters.  
\begin{equation}
    G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right) = \mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right).
\end{equation}
This model is completed with a normal prior on $\bm{\mu}$, an inverse Wishart prior on $\bm{\Sigma}$, and Gamma priors on $\zeta_{\ell}$ and $\sigma_{\ell}$, and $\eta$.  This model is denoted as the \emph{projected Gamma--log-normal} (PG--LN) model.  We also explore a restricted Gamma form of this model as above, where $\beta_{\ell} = 1$ for all $\ell$.  This is denoted as the \emph{projected restricted Gamma--log-normal} (PRG--LN) model.  In the interest of brevity, we omit the inferential forms for this model, sufficing to say that the Gamma prior for $\beta_{j\ell}$ means it can again be integrated out, improving mixing on sampling $\bm{\alpha}_j$ with a joint Metropolis step; the normal prior for $\mu$ is conjugate for the log-normal $\bm{\alpha}_j$, and sampled via a Gibbs step; and an inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, implying that it can also be sampled via a Gibbs step.
